### 2023-12-26 sections/4-results.tex
more additions to discussion as well as fixing some inconsistencies in results

Figures~\ref{fig:results:fashion-mnist-scaling}, ~\ref{fig:results:glove-25-scaling}, and ~\ref{fig:results:sift-scaling} show the scaling behavior of CAKES algorithms and existing algorithms on augmented versions of the following ann-benchmark datasets: fashion-mnist under Euclidean distance, glove-25 under cosine distance, and sift under Euclidean distance.
Figure~\ref{fig:results:random-scaling} shows the scaling behavior of CAKES algorithms and existing algorithms on a completely randomly generated dataset with same cardinality and dimensionality as sift (1 million points in 128 dimensions).
Figures~\ref{fig:results:silva-scaling} and ~\ref{fig:results:radioml-scaling} show the scaling behavior of CAKES on Silva and RadioML respectively.
The horizontal axis in each figure shows the cardinality of the augmented dataset with synthetic points (see Section \ref{subsec:methods:synthetic-data}).
The left-most point on each line is on the original dataset without any synthetic augmentation.
The vertical axis denotes throughput in queries per second.
Both axes are on a logarithmic scale.
In this section, we report results only for $k$-NN search with $k = 10$, but similar plots for $k=100$ can be found in the Supplement.
In these plots, we label the recall for algorithms which do not exhibit perfect recall;
if no such labels are given for an algorithm, then that algorithm exhibits perfect recall.

Tables \ref{table:results:ann-fashion}, \ref{table:results:ann-glove-25}, \ref{table:results:ann-sift}, and \ref{table:results:ann-random} show the throughput and recall of CAKES's algorithms at each cardinality multiplier for the fashion-mnist, glove-25, sift, and random datasets respectively.
When reporting recall, we use 1.000* to denote that the recall is imperfect, but rounds to 1.000 when we consider only three decimal places.

In Figure~\ref{fig:results:fashion-mnist-scaling}, which shows results for fashion-mnist, we observe that as the multiplier increases, the CAKES algorithms (GreedySieve in blue, RepeatedRnn in green, and Sieve in red) exhibit higher throughput than na\"{i}ve linear search, plotted in orange.
We also observe that CAKES outperforms FAISS-IVF (plotted in pink) for all multipliers greater than one.
Though state-of-the-art algorithms HNSW (plotted in gray), and ANNOY (plotted in purple) have higher throughput than CAKES for all cardinalities, we note that CAKES always exhibits perfect recall while HNSW and ANNOY exhibit much lower recall, as shown in Table~\ref{table:results:ann-fashion}.
For example, at a multiplier as low as eight, HNSW and ANNOY have recall of 0.525 and 0.857 respectively,
and this recall gets worse as the multiplier increases.
Meanwhile, while CAKES exhibits perfect recall.
Amongst the CAKES algorithms, GreedySieve consistently performs best on this dataset, exhibiting sublinear time performance with increasing multipliers.

With glove-25 (Figure~\ref{fig:results:glove-25-scaling}), we observe that each of CAKES' algorithms outperforms linear search at high multipliers.
Similar to the results for fashion-mnist, we see that CAKES outperforms FAISS-IVF for all multipliers greater than four.
We see once again that HNSW and ANNOY have higher throughput than CAKES for all multipliers, CAKES exhibits better recall at all multipliers, as shown in Table~\ref{table:results:ann-glove-25}.
In particular, at a multiplier as low as two, HNSW and ANNOY have recall as low as 0.607 and 0.832 respectively, and this recall continues to decrease as we increase the multiplier.
In contrast, CAKES exhibits perfect recall at all multipliers.
On this dataset, unlike with fashion-mnist, Repeated $\rho$-NN, plotted in green, is consistently the best-performing CAKES algorithm.

In Figure~\ref{fig:results:sift-scaling}, which displays the scaling behavior of algorithms on the sift dataset, we observe that while linear search performs best for low multipliers, our algorithms begin to outperform linear search at a multiplier of eight.
Again, we observe that CAKES outperforms FAISS-IVF for all multipliers greater than four. 
For this dataset, Sieve, plotted in red, is the highest throughput CAKES algorithm for all multipliers greater than right.
Similar to the results with fashion-mnist and glove-25, we observe that while HNSW and Annoy exhibit higher throughput than CAKES for all multipliers, CAKES has better recall.
Table~\ref{table:results:ann-sift} shows that even with a multiplier of 1 (i.e., the original dataset), HNSW and Annoy have recall as low as 0.782 and 0.686 respectively, while CAKES exhibits perfect recall.

With the random dataset, as seen in Figure~\ref{fig:results:random-scaling}, we observe that the three CAKES algorithms exhibit the worst throughput out of all algorithms, at all multipliers. 
As with the real datasets, HNSW and ANNOY exhibit the highest throughput of any of the algorithms and CAKES exhibits perfect recall at all multipliers. 
HNSW and Annoy exhibit \emph{much} lower recall on this random dataset than on the real datasets; 
in particular, with a multiplier of 1, HNSW and Annoy have recall as low as 0.060 and 0.028 respectively, as reported in Table~\ref{table:results:ann-random}.

@@

\begin{table*}[!t]
    % \renewcommand{\arraystretch}{1.15}
    \caption{Runtime performance (queries per second) and recall on the \textbf{glove-25} dataset. 1.000* denotes imperfect recall that rounds to 1.000.}
    \label{table:results:ann-glove-25}
    \vskip 0.15in
    \begin{center}
        \begin{small}
            \begin{sc}
                \begin{tabular}{|l|p{1.2cm}|p{0.8cm}|p{1.2cm}|p{0.8cm}|p{1.2cm}|p{0.8cm}|p{1.2cm}|p{0.8cm}|p{1.2cm}|p{0.8cm}|}
                    \hline
                    \textbf{Scale}  & \multicolumn{2}{|c|}{\textbf{hnsw}} & \multicolumn{2}{|c|}{\textbf{annoy}} & \multicolumn{2}{|c|}{\textbf{faiss-flat}} & \multicolumn{2}{|c|}{\textbf{faiss-ivf}}  & \multicolumn{2}{|c|}{\textbf{CAKES}} \\
                    \hline
                    &             QPS & Recall        & QPS & Recall      & QPS & Recall       & QPS & Recall     & QPS & Recall    \\
                    \hline
                    1   & 2.28~$\times10^{4}$ & 0.801 & 2.83~$\times10^{3}$ & 0.835 & 2.78~$\times10^{2}$ & 1.000 & 2.38~$\times10^{3}$ & 1.000* & 7.22~$\times10^{2}$ & 1.000 \\
                    \hline
                    2   & 2.38~$\times10^{4}$ & 0.607 & 2.70~$\times10^{3}$ & 0.832 & 1.39~$\times10^{2}$ & 1.000 & 1.19~$\times10^{3}$ & 1.000* & 5.75~$\times10^{2}$ & 1.000 \\
                    \hline
                    4   & 2.50~$\times10^{4}$ & 0.443 & 2.61~$\times10^{3}$ & 0.839 & 6.98~$\times10^{1}$ & 1.000 & 6.19~$\times10^{2}$ & 1.000* & 6.25~$\times10^{2}$ & 1.000 \\
                    \hline
                    8   & 2.78~$\times10^{4}$ & 0.294 & 2.51~$\times10^{3}$ & 0.834 & 3.49~$\times10^{1}$ & 1.000 & 3.03~$\times10^{2}$ & 1.000* & 5.93~$\times10^{2}$ & 1.000 \\
                    \hline
                    16  & 3.11~$\times10^{4}$ & 0.213 & 2.23~$\times10^{3}$ & 0.885 & 1.74~$\times10^{1}$ & 1.000 & 1.51~$\times10^{2}$ & 1.000* & 5.49~$\times10^{2}$ & 1.000 \\
                    \hline
                    32  & 3.24~$\times10^{4}$ & 0.178 & 2.01~$\times10^{3}$ & 0.764 & 8.70~$\times10^{0}$ & 1.000 & 7.40~$\times10^{1}$ & 0.999  & 4.75~$\times10^{2}$ & 1.000 \\
                    \hline
                    64  & --                  & --    & 1.99~$\times10^{3}$ & 0.631 & 4.36~$\times10^{0}$ & 1.000 & 3.77~$\times10^{1}$ & 0.997  & 4.61~$\times10^{2}$ & 1.000 \\
                    \hline
                    128 & --                  & --    & --                  & --    & 2.18~$\times10^{0}$ & 1.000 & 1.90~$\times10^{1}$ & 0.998  & 4.41~$\times10^{2}$ & 1.000 \\
                    \hline
                    256 & --                  & --    & --                  & --    & 1.07~$\times10^{0}$ & 1.000 & 9.47~$\times10^{0}$ & 0.998  & 4.23~$\times10^{2}$ & 1.000 \\
                    \hline
                    % 512 & -- & -- & -- & -- & 0.544 & 1.000 & 4.499 & 0.996 & 0.000 & 0.000 \\
                    % \hline
                \end{tabular}
            \end{sc}
        \end{small}
    \end{center}
    \vskip -0.1in
\end{table*}

>>>>

Figures~\ref{fig:results:fashion-mnist-scaling}, ~\ref{fig:results:glove-25-scaling}, and ~\ref{fig:results:sift-scaling} show the scaling behavior of CAKES algorithms and existing algorithms on augmented versions of the following ann-benchmark datasets: fashion-mnist under Euclidean distance, glove-25 under cosine distance, and sift under Euclidean distance.
Figure~\ref{fig:results:random-scaling} shows the scaling behavior of CAKES algorithms and existing algorithms on a completely randomly generated dataset with same cardinality and dimensionality as sift (1 million points in 128 dimensions).
Figures~\ref{fig:results:silva-scaling} and ~\ref{fig:results:radioml-scaling} show the scaling behavior of CAKES on Silva and RadioML respectively.
The horizontal axis in each figure shows the cardinality of the augmented dataset with synthetic points (see Section \ref{subsec:methods:synthetic-data}).
The left-most point on each line is on the original dataset without any synthetic augmentation.
The vertical axis denotes throughput in queries per second.
Both axes are on a logarithmic scale.
In this section, we report results only for $k$-NN search with $k = 10$, but similar plots for $k=100$ can be found in the Supplement.
{\color{red} TODO: revisit this sentence with cosine dist plot in mind: In these plots, we label the recall for algorithms which do not exhibit perfect recall; if no such labels are given for an algorithm, then that algorithm exhibits perfect recall" }.

Tables \ref{table:results:ann-fashion}, \ref{table:results:ann-glove-25}, \ref{table:results:ann-sift}, and \ref{table:results:ann-random} show the throughput and recall of CAKES's algorithms at each cardinality for  Fashion-mnist, Glove-25, Sift, and a random dataset respectively. Though the scaling plots above present results for each of CAKES's three algorithms separately, the results in the CAKES column in these tables represent the highest throughput CAKES algorithm at that dataset and cardinality only. 
Given that we redo the auto-tuning process described in Section~\ref{subsec:methods:auto-tuning} for each augmented dataset, these values are an accurate representation of CAKES's performance on each dataset and cardinality multiplier. 
When reporting recall, we use 1.000* to denote that the recall is imperfect, but rounds to 1.000 when we consider only three decimal places.

In Figure~\ref{fig:results:fashion-mnist-scaling}, which shows results for Fashion-mnist, we observe that as cardinality increases, the CAKES algorithms (GreedySieve in blue, Repeated $\rho$-NN in green, and Sieve in red) exhibit higher throughput than na\"{i}ve linear search (plotted in orange).
In particular, we see that GreedySieve and Sieve begin to exhibit sublinear time performance at a cardinality near $10^5$ (i.e., at a multiplier of about {\color{red} add value} times the original dataset's cardinality). 
We also observe that CAKES's algorithms outperform FAISS-IVF (plotted in pink) for all cardinalities larger than the original dataset's cardinality.
Though state-of-the-art algorithms HNSW (plotted in gray), and Annoy (plotted in purple) have higher throughput than CAKES for all cardinalities, we note that CAKES always exhibits perfect recall while HNSW and Annoy exhibit much lower recall, as shown in Table~\ref{table:results:ann-fashion}.
For example, at a multiplier as low as eight, HNSW and ANNOY have recall of 0.525 and 0.857 respectively,
and this recall gets worse as the multiplier increases.
Meanwhile, CAKES exhibits perfect recall.
Amongst the CAKES algorithms, GreedySieve consistently performs best on this dataset, exhibiting sublinear time performance with increasing multipliers.

With glove-25 (Figure~\ref{fig:results:glove-25-scaling}), we observe that each of CAKES' algorithms outperforms linear search at some cardinality.
Specifically, all three of CAKES' algorithms outperform linear search at a cardinality just slightly more than the original dataset's cardinality, $10^6$.
Similar to the results for Fashion-mnist, we see that CAKES's algorithms outperform FAISS-IVF for all cardinalities greater than $10^7$.
We see once again that HNSW and Annoy have higher throughput than CAKES's algorithms for all cardinality multipliers, CAKES's algorithms exhibit better recall at all multipliers, as shown in Table~\ref{table:results:ann-glove-25}.
In particular, at a multiplier as low as two, HNSW and ANNOY have recall as low as 0.607 and 0.832 respectively, and this recall continues to decrease as we increase the multiplier.
In contrast, CAKES exhibits perfect recall at all multipliers.
On this dataset, unlike with fashion-mnist, Repeated $\rho$-NN, plotted in green, is consistently the best-performing CAKES algorithm.

In Figure~\ref{fig:results:sift-scaling}, which displays the scaling behavior of algorithms on the Sift dataset, we observe that while linear search performs best for lower cardinalities, our algorithms begin to outperform our implementation of linear search and FAISS-IVF at a cardinality near $10^7$.
For this dataset, Sieve, plotted in red, is the highest-throughput CAKES algorithm for all cardinalities.
Similar to the results with Fashion-mnist and Glove-25, we observe that while HNSW and Annoy exhibit higher throughput than CAKES's algorithms for all cardinality multipliers, CAKES has better recall.
Table~\ref{table:results:ann-sift} shows that even with a multiplier of 1 (i.e., the original dataset), HNSW and Annoy have recall as low as 0.782 and 0.686 respectively, while CAKES exhibits perfect recall.

With the random dataset, as seen in Figure~\ref{fig:results:random-scaling}, we observe that the three CAKES algorithms exhibit the worst throughput out of all algorithms, at all cardinalities. 
As with the real datasets, HNSW and Annoy exhibit the highest throughput of any of the algorithms, and CAKES exhibits perfect recall at all cardinalities. 
HNSW and Annoy exhibit \emph{much} lower recall on this random dataset than on the real datasets; 
in particular, with a multiplier of 1, HNSW and Annoy have recall as low as 0.060 and 0.028 respectively, as reported in Table~\ref{table:results:ann-random}.

@@

\begin{table*}[!t]
    % \renewcommand{\arraystretch}{1.15}
    \caption{Runtime performance (queries per second) and recall on the \textbf{glove-25} dataset. 1.000* denotes imperfect recall that rounds to 1.000.}
    \label{table:results:ann-glove-25}
    \vskip 0.15in
    \begin{center}
        \begin{small}
            \begin{sc}
                \begin{tabular}{|l|p{1.2cm}|p{0.8cm}|p{1.2cm}|p{0.8cm}|p{1.2cm}|p{0.8cm}|p{1.2cm}|p{0.8cm}|p{1.2cm}|p{0.8cm}|}
                    \hline
                    \textbf{Scale}  & \multicolumn{2}{|c|}{\textbf{hnsw}} & \multicolumn{2}{|c|}{\textbf{annoy}} & \multicolumn{2}{|c|}{\textbf{faiss-flat}} & \multicolumn{2}{|c|}{\textbf{faiss-ivf}}  & \multicolumn{2}{|c|}{\textbf{CAKES}} \\
                    \hline
                    &             QPS & Recall        & QPS & Recall      & QPS & Recall       & QPS & Recall     & QPS & Recall    \\
                    \hline
                    1   & 2.28~$\times10^{4}$ & 0.801 & 2.83~$\times10^{3}$ & 0.835 & 2.78~$\times10^{2}$ & 1.000 & 2.38~$\times10^{3}$ & 1.000* & 7.22~$\times10^{2}$ & 1.000* \\
                    \hline
                    2   & 2.38~$\times10^{4}$ & 0.607 & 2.70~$\times10^{3}$ & 0.832 & 1.39~$\times10^{2}$ & 1.000 & 1.19~$\times10^{3}$ & 1.000* & 5.75~$\times10^{2}$ & 1.000* \\
                    \hline
                    4   & 2.50~$\times10^{4}$ & 0.443 & 2.61~$\times10^{3}$ & 0.839 & 6.98~$\times10^{1}$ & 1.000 & 6.19~$\times10^{2}$ & 1.000* & 6.25~$\times10^{2}$ & 1.000* \\
                    \hline
                    8   & 2.78~$\times10^{4}$ & 0.294 & 2.51~$\times10^{3}$ & 0.834 & 3.49~$\times10^{1}$ & 1.000 & 3.03~$\times10^{2}$ & 1.000* & 5.93~$\times10^{2}$ & 1.000* \\
                    \hline
                    16  & 3.11~$\times10^{4}$ & 0.213 & 2.23~$\times10^{3}$ & 0.885 & 1.74~$\times10^{1}$ & 1.000 & 1.51~$\times10^{2}$ & 1.000* & 5.49~$\times10^{2}$ & 1.000* \\
                    \hline
                    32  & 3.24~$\times10^{4}$ & 0.178 & 2.01~$\times10^{3}$ & 0.764 & 8.70~$\times10^{0}$ & 1.000 & 7.40~$\times10^{1}$ & 0.999  & 4.75~$\times10^{2}$ & 1.000* \\
                    \hline
                    64  & --                  & --    & 1.99~$\times10^{3}$ & 0.631 & 4.36~$\times10^{0}$ & 1.000 & 3.77~$\times10^{1}$ & 0.997  & 4.61~$\times10^{2}$ & 1.000* \\
                    \hline
                    128 & --                  & --    & --                  & --    & 2.18~$\times10^{0}$ & 1.000 & 1.90~$\times10^{1}$ & 0.998  & 4.41~$\times10^{2}$ & 1.000* \\
                    \hline
                    256 & --                  & --    & --                  & --    & 1.07~$\times10^{0}$ & 1.000 & 9.47~$\times10^{0}$ & 0.998  & 4.23~$\times10^{2}$ & 1.000* \\
                    \hline
                    % 512 & -- & -- & -- & -- & 0.544 & 1.000 & 4.499 & 0.996 & 0.000 & 0.000 \\
                    % \hline
                \end{tabular}
            \end{sc}
        \end{small}
    \end{center}
    \vskip -0.1in
\end{table*}

### 2023-12-26 sections/5-discussion.tex
more additions to discussion as well as fixing some inconsistencies in results

We have presented CAKES, a method for fast $k$-nearest-neighbors search that can adapt to a variety of distance functions.
CAKES provides exact recall with $k$-NN when the distance function in use is a metric (particularly, when it obeys the triangle inequality).
CAKES is most effective when datasets are not uniform and do not obey a ``simple'' distribution such as Gaussian, but rather adhere to ``interesting'' low-dimensional manifolds even when embedded in a high-dimensional space;
thus, CAKES works best when the manifold hypothesis is upheld.

In comparison to other methods, while CAKES has perfect recall for distance metrics, it exhibits worse recall under non-metrics (specifically, cosine distance). This bears further investigation, but likely stems from how heavily the $k$-NN and clustering algorithms depend on the triangle inequality when discarding clusters from consideration.

>>>>

We have presented CAKES, a set of three algorithms for fast $k$-nearest-neighbors search that can adapt to a variety of distance functions.
CAKES's algorithms are exact when the distance function in use is a metric (particularly, when it obeys the triangle inequality). When we benchmarked the algorithms with cosine distance, a non-metric, we found that CAKES has \emph{nearly} perfect recall; at three decimal places of precision, CAKES's algorithms have recall indistinguishable from perfect.

CAKES's algorithms were designed to be most effective when the dataset is not uniformly distributed, but rather adheres to a low-dimensional manifold even when embedded in a high-dimensional space; 
in other words, CAKES was designed to work best when the manifold hypothesis is upheld.Our results indicate that CAKES algorithms do indeed perform well on datasets that adhere to the manifold hypothesis, and that they perform poorly on datasets that do not. 
We use the LFD by depth percentiles shown in Section~\ref{subsec:lfd-results} to quantify the extent to which a dataset obeys the manifold hypothesis; we say that
the lower a dataset's LFD, the more that dataset exhibits a manifold structure. We observe a striking example of this difference when contrasting CAKES's algorithms' performance on Sift versus on the random dataset with the same cardinality and dimensionality as Sift.

As seen in Section~\ref{subsec:lfd-results}, the local fractal dimension of the random dataset is much higher than that of Sift.
This is unsurprising, given that the random dataset is, by definition, uniformly distributed, while Sift is not.
On the random dataset, CAKES's algorithms' throughputs decrease as the cardinality multiplier increases, while with Sift, GreedySieve and Sieve both exhibit nearly constant throughput.
While Repeated $\rho$-NN does not exhibit constant throughput with Sift, it does exhibit a much slower decrease in throughput than it does with the random dataset.
Since these two datasets are the same size, the aforementioned discrepancies isolate the effect of the manifold structure of the dataset on algorithm performance.
We emphasize that though CAKES's algorithms' throughputs are lower on a random dataset, their recalls remain perfect at all multipliers for both datasets. 
A sort of converse of this trend occurs with HNSW and Annoy. 
In contrast, for these two algorithms, \emph{throughput} remains constant with cardinality multiplier on both datasets, but their \emph{recall} for the random dataset is orders of magnitude worse than their recall for Sift.
These results illustrate the difference between CAKES's algorithms and the state-of-the-art similarity search algorithms. They also demonstrate CAKES's algorithms suitability on the type of datasets for which they were designed.

We also note that on different datasets which exhibit manifold structures, the highest-throughput CAKES algorithm varies. 
For example, on Glove-25, Repeated $\rho$-NN exhibits the highest throughput of any CAKES algorithm, despite having the lowest throughput of the three on Sift and Fashion-mnist. 
On Silva,  Repeated $\rho$-NN is comparable to the other CAKES algorithms at low multipliers, but becomes the best of the three at higher multipliers. 
When we view these results in light of the LFD plots in Section~\ref{subsec:lfd-results}, we realize that  Repeated $\rho$-NN appears to be the best-performing CAKES algorithm on datasets where a large proportion of the dataset has a very low LFD (i.e., near 2), like Glove-25 and Silva. 
This general trend is unsurprising, given that Repeated $\rho$-NN relies on a low LFD around the query in order to quickly home in on the correct radius for $k$ hits. 
While it varies which of the two algorithms has better throughput, we observe that on all three ANN benchmark datasets (Fashion-mnist, Glove-25, and Sift), GreedySieve's and Sieve's throughput stays nearly constant as the cardinality multiplier increases. 
This observation warrants further investigation, but it is especially promising given that GreedySieve and Sieve consistently outperform linear search for high cardinalities. 
Finally, we emphasize that the variation in performance of our algorithms across different datasets and cardinalities support our use of an auto-tuning function (as discussed in Section~ref{subsec:methods:auto-tuning}) to select the best choice of algorithm for a given dataset. 
Given the trends such as that with low LFD and Repeated $\rho$-NN's performance, we suspect that a more sophisticated, less-heuristic auto-tuning function could be developed to select the best algorithm based on the properties of the dataset.

We also stress that the CAKES's algorithms are designed to work well for \emph{big} data, and our results support this claim; we observe that while linear search can outperform CAKES on datasets with small cardinality, CAKES always overtakes linear search at some cardinality multiplier for each of the ANN benchmark datasets we tested on. Moreover, we notice that the cardinality multiplier at which CAKES overtakes linear search is lower for datasets with lower LFD. For example, Sift has much higher LFD than Fashion-mnist and Glove-25, and we observe that CAKES overtakes linear search at a cardinality of $10^7$ for Sift, as opposed to about $10^5$ and $10^6$ for Fashion-mnist and Glove-25, respectively. For the truly random dataset, which had the highest LFD, CAKES's algorithms \emph{never} outperform linear search. These observations support our claims that CAKES's algorithms work well with big datasets, and further support our claims that they are much better suited to datasets which exhibit a manifold structure.
